---
title: "DATA 608 Assignment 1"
author: "Nathaniel Cooper Ph.D."
date: "August 27, 2018"
output:
  html_document: default
  html_notebook: default
---

**Principles of Data Visualization and Introduction to ggplot2**

I have provided you with data about the 5,000 fastest growing companies in the US, as compiled by Inc. magazine. lets read this in:

```{r}

inc <- read.csv("https://raw.githubusercontent.com/charleyferrari/CUNY_DATA_608/master/module1/Data/inc5000_data.csv", header= TRUE)

```

And lets preview this data:

```{r}

head(inc)

```
```{r}

summary(inc)

```

Think a bit on what these summaries mean. Use the space below to add some more relevant non-visual exploratory information you think helps you understand this data:

```{r}
suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library("RColorBrewer")))
```

```{r}
#Summary doesn't include Standard Deviation so we can include that on the numerical data:
sd(inc$Growth_Rate)
sd(inc$Revenue)
sd(inc$Employees, na.rm = TRUE)#A few companies have missing employee counts

#We can also do IQR in case the data is skewed

IQR(inc$Growth_Rate)
IQR(inc$Revenue)
IQR(inc$Employees, na.rm = TRUE)

#Revenue has quite a large range. I used a base 10 logrithm to compress that scale.
#I used mutate from tidyr to make these new calculations part of the dataframe for later use.
inc <- inc %>% mutate(log_Rev = log10(Revenue))
inc$log_Rev %>% summary()

# I also like doing ratios, in this case normalizing businesses of different sizes to see how much revenue or growth is generated per employee. 
inc <- inc %>% mutate(rev_per_empl = Revenue/Employees)
inc$rev_per_empl %>% summary() 

inc <- inc %>% mutate(grw_per_empl = Growth_Rate/Employees)
inc$grw_per_empl %>% summary()

inc <- inc %>% mutate(log_rev_per_grw = log10(Revenue/Growth_Rate))
inc$log_rev_per_grw %>% summary()
```

## Question 1

Create a graph that shows the distribution of companies in the dataset by State (ie how many are in each state). There are a lot of States, so consider which axis you should use. This visualization is ultimately going to be consumed on a 'portrait' oriented screen (ie taller than wide), which should further guide your layout choices.

For all three questions, we are asked to give distributions of catagorical data: States or Industries. The simple Bar Chart seems to be the most intuitive way to present the information. I used horizontal bars to make the charts fit the portrait orientation better.

```{r}
state_counts <- inc$State %>% table() %>% as.data.frame(stringsAsFactors
=FALSE)
colnames(state_counts) <- c('State', 'Frequency')


ggplot(state_counts, aes(x=reorder(State, Frequency),y=Frequency, color=State)) +
  geom_bar(stat='identity', color = 'black', fill=rainbow(52)) +
  coord_flip() + 
  xlab('US State')+
  theme_classic()
#Ordering adapted from: https://stackoverflow.com/questions/25664007/reorder-bars-in-geom-bar-ggplot2
```


I wanted to include every state, so the bars are very thin. I used a color gradient with the rainbow pallette to help people distinguish the bars. This does reduce the data-ink ratio quite a bit, but I think, makes the graph far easier to understand.

## Quesiton 2

Lets dig in on the state with the 3rd most companies in the data set. Imagine you work for the state and are interested in how many people are employed by companies in different industries. Create a plot that shows the average and/or median employment by industry for companies in this state (only use cases with full data, use R's `complete.cases()` function.) In addition to this, your graph should show how variable the ranges are, and you should deal with outliers.

From the above chart, New York is in third place for high growth companies. I simply used tidyr's filter() function to subset the data. I used error bars to show the range of the variables. For the industries that have a single entry, and therefore undefined standard deviation, I imputed the value of 0 to draw the error bars.

```{r}
#apadted: https://stackoverflow.com/questions/28687515/search-for-and-remove-outliers-from-a-dataframe-grouped-by-a-variable
ny_data <- inc  %>% filter(State == 'NY', complete.cases(.)) %>% arrange(Industry) %>% select(Industry, Employees) 
ny_data <- ny_data %>% group_by(Industry) %>% filter(!(abs(Employees - median(Employees)) > 1.5*IQR(Employees)))# Using 1.5xIQR as the outlier limit
ny_data
#The 1.5xIQR rule reduced the number of negative error bars better than the 2xstd dev rule.

ind_means <- ny_data %>% group_by(Industry) %>% summarise(mean_emp = mean(Employees), emp_sd = sd(Employees))
ind_means$emp_sd[is.na(ind_means$emp_sd)] <- 0


ind_means


ggplot(ind_means, aes(x=reorder(Industry, mean_emp),y=mean_emp)) +
  geom_bar(stat='identity', color = 'black', fill='lightgray') +
  geom_errorbar(aes(ymin = mean_emp - emp_sd, ymax = mean_emp + emp_sd), width=0.2) +
  theme(legend.position="none") +
  ylab('Mean Employees')+ xlab('Industry')+ 
  coord_flip() +
  theme_classic()
```

Since there are fewer bars, but the enough where spacing between them is small, I used light gray with a dark border to help distinguish the bars. This increases the data-ink ratio from the above bar plot.


## Question 3

Now imagine you work for an investor and want to see which industries generate the most revenue per employee. Create a chart that makes this information clear. Once again, the distribution per industry should be shown.

I had previously calculated the ratios for revenue per employee. I used the same 1.5$\times$IQR as the outlier range and repeated the process above to make the bar chart.

```{r}
rev_per_ind <- inc %>% filter(complete.cases(.)) %>% group_by(Industry) %>% filter(!(abs(rev_per_empl - median(rev_per_empl)) > 1.5*IQR(rev_per_empl))) %>% summarise(mean_rev_per_empl = mean(rev_per_empl), rev_sd = sd(rev_per_empl))
rev_per_ind

ggplot(rev_per_ind , aes(x=reorder(Industry,mean_rev_per_empl),y=mean_rev_per_empl)) +
  geom_bar(stat='identity', color = 'black', fill='lightgray') +
  geom_errorbar(aes(ymin = mean_rev_per_empl - rev_sd, ymax = mean_rev_per_empl + rev_sd), width=0.2) +
  theme(legend.position="none") +
  ylab('Mean Revenue per Employee')+ xlab('Industry')+
  coord_flip()+
  theme_classic()
```

I used the same design philosophy as number 2.
